---
title: "Goose Migration Sept 2022"
author: "Shannon Curley and Michael Allen"
date: '2023-10-27'
output: html_document


```{r}
### Libraries
library(dplyr) ### Data manipulation
library(tidyverse) #dataframe manipulation and to remove NAs from columns

########################################################
### Filtering the data
########################################################


### Load USGS data--> File sent directly from Danny Bystrack from request
Curley_Ramirez_Geese_Es_202008201701 <- read.csv("~/Desktop/Goose_Ecography/Goose_Final/Data/Curley_Ramirez_Geese_Es_202008201701.csv")#1,611,483 obs

### Grab the needed columns 
Goose_data<-dplyr::select(Curley_Ramirez_Geese_Es_202008201701,B_COORD_PRECISION,B_COUNTRY_CODE, B_LAT_DECIMAL_DEGREES, B_LON_DECIMAL_DEGREES, B_SPECIES_ID, B_SPECIES_NAME, BAND_NUM, BANDING_DAY, BANDING_MONTH, BANDING_YEAR, BEARING, DISTANCE, E_COORD_PRECISION, E_COUNTRY_CODE, E_LAT_DECIMAL_DEGREES, E_LON_DECIMAL_DEGREES, ENCOUNTER_MONTH, ENCOUNTER_YEAR, SAME_10_MIN_BLOCK)

### Clean
rm(Curley_Ramirez_Geese_Es_202008201701)

###Grab the species that we want - removed: All Canada Goose species, Barnacle Goose (9 records) Emporer Goose (201), Integrades (Black X Atlantic Brant Intergrade), (Snow X Blue Goose Intergrade ), (Hyrids, Snow X Ross's Goose Hybrid,  Tule White-fronted Goose, ), And "Blue Geese"
Goose_data<-Goose_data %>% filter(B_SPECIES_NAME %in% c("Lesser Snow Goose","Ross's Goose","Greater White-fronted Goose","Cackling Goose","Atlantic Brant","Black Brant", "Greater Snow Goose")) #240,654 obs

###Filter Years (data goes back to 1950)
Goose_data<-Goose_data%>% filter(BANDING_YEAR >= 1990) #140,443 obs
Goose_data<-Goose_data%>% filter(BANDING_YEAR <= 2019) #140,390 obs

###Remove encounter year 2020 because that year was only midway compiled when we requested the data
Goose_data<-Goose_data%>% filter(ENCOUNTER_YEAR <= 2019) #136,452 obs

###Filter months for winter and breeding season
Goose_data<-Goose_data %>% filter(BANDING_MONTH %in% (5:7)) #64,536 obs

Goose_data<-Goose_data%>% filter(ENCOUNTER_MONTH %in% c(12, 1, 2)) #31,472 obs

### Remove species recorded in same 10 block (to remove bias of birds that did not migrate)
Goose_data<-Goose_data%>% filter(SAME_10_MIN_BLOCK %in% "N") #31,468 obs

### Remove data with no distance information (these are usually data from hunters without exact dates)
Goose_data<-Goose_data %>% drop_na(DISTANCE) #31,409 obs

### Remove duplicates (occurs when 2 different people record the same bird at the same time)
Goose_data<-distinct(Goose_data)#31,371 obs

### Remove breeding records below 50 degrees latitude (retain arctic and subarctic records)
Goose_data<-Goose_data%>% filter(B_LAT_DECIMAL_DEGREES >= 50) #30,797 obs

### Only retain birds banded in US and CA
b_country_code<-c("CA", "US")
Goose_data<-Goose_data%>% filter(B_COUNTRY_CODE %in% b_country_code) #28,665 obs

### and remove birds encounter outside of continental N and S america (rm Russa and Japan)
table(Goose_data$E_COUNTRY_CODE)

e_country_code<-c("CA", "US", "MX")
Goose_data<-Goose_data%>% filter(E_COUNTRY_CODE %in% e_country_code) #28,656 obs

### Remove encounter coordinates
## No need to remove Banding precision but double check
#table(Goose_data$B_COORD_PRECISION)
prec_values<-c(0, 1, 10, 11) # categories with higher coordinate resoltuion
Goose_data<-Goose_data%>% filter(E_COORD_PRECISION %in% prec_values) #28,085 obs
### see https://www.pwrc.usgs.gov/BBL/manual/inexact.cfm for definitions

### Check for birds that have more than one record to make sure they don't repeat
more_than_2<-as.data.frame(table(Goose_data$BAND_NUM))
more_than_2<-subset(more_than_2, subset = more_than_2$Freq>=2) #194 bands repeated 2-5 times

### Make a df pulling just these repeated band numbers to filter, but also remove them from Goose df

### Remove these bands from the original dataset and will return once filtered and examined
more_than_2_records<-Goose_data %>%
  filter(BAND_NUM %in% more_than_2$Var1) #431 obs to be examined and filtered

### Remove these bands from the original dataset and will return once filtered and examined
Goose_data<-Goose_data %>%
  filter(!BAND_NUM %in% more_than_2$Var1) #27,654 obs

### Check that the difference in goose df matches 565 records to be filtered
#28,085-431=27654 YES

repeat_bands<-unique(more_than_2_records$BAND_NUM)

repeat_df<-data.frame()

### for birds with values more than once in a year, grab the distance traveled the farthest - removes duplicate birds, however, we then need to see if birds were encountered in dec 1999 and then in jan or feb of following year (2000) 

for (i in 1:(length(repeat_bands))){
  
by_band<-dplyr::filter(more_than_2_records, BAND_NUM==repeat_bands[i])

years<-sort(unique(by_band$ENCOUNTER_YEAR))

for (j in 1:(length(years))){

by_year<-dplyr::filter(by_band, ENCOUNTER_YEAR==years[j])

to_save<-by_year[which.max(by_year$ENCOUNTER_YEAR),]


repeat_df<-rbind(repeat_df, to_save)


}
}

### repeat df removes 64 individuals that were recorded in the same year, and retained the max distance traveled.


### import visually inspected df 
repeat_df_3 <- read.csv("~/Desktop/Goose_Ecography/Goose_revisions/repeat_df_3.csv")

Goose_data<-rbind(Goose_data, repeat_df_3)

write.csv(Goose_data, file="Goose_data_final_OCT2023.csv", row.names = F)


Goose_data<-filter(Goose_data, B_SPECIES_NAME==!c("Greater Snow Goose"))

```


---
# Load libraries; load & format final data
```{r}

########################################################
### Bayesian Models
########################################################


library(dplyr) ### Data manipulation
library(ggplot2)
'%notin%' <- Negate('%in%')

# import data #make sure plyr is not installed
FINAL_GOOSE_DATA <- f <- read.csv("~/Desktop/Goose_Ecography/Goose_revisions/Goose_data_final_OCT2023.csv") %>%
  rename(dist = DISTANCE,
         year = ENCOUNTER_YEAR,
         species = B_SPECIES_NAME,
         lat = B_LAT_DECIMAL_DEGREES,
         lon = B_LON_DECIMAL_DEGREES
         ) %>%
  mutate(year = year - 1989, 
         int = year*lat,
         zy = wiqid::standardize(dist),
         zx = wiqid::standardize(year),
         zlat = wiqid::standardize(lat),
         zlon = wiqid::standardize(lon),
         zint = wiqid::standardize(int))

# format data for JAGS
goose_jags_data <- list(
  y = f$dist,
  x = as.numeric(f$year),
  s = as.numeric(as.factor(f$species)),
  Nsubj = length(unique(f$species)),
  lat=f$lat,
  lon=f$lon,
  int=f$int,
  zy = f$zy,
  zx = f$zx,
  zlat = f$zlat,
  zlon = f$zlon,
  zint = f$zint,
  Ntotal = length(f$dist)
)
```
# more data exploration with latitude
```{r}
library(ggplot2)
ggplot(f) +
  geom_point(aes(x = lat, y = dist, color = year)) +
  facet_wrap(~species) +
  viridis::scale_color_viridis(option = "inferno")

ggsave("dist_v_lat.png", height = 6, width = 8, dpi = 400)

```
# Specify Bayesian model - year + latitude + year:latitude 
Info about how to deal with interaction terms when standardizing variables...
i.e., with standardized product of variables, not product of standardized variables
https://stats.stackexchange.com/questions/151468/adding-interaction-terms-to-multiple-linear-regression-how-to-standardize
https://www.statalist.org/forums/forum/general-stata-discussion/general/1501450-standardizing-an-interaction-term
Note: JAGS code adapted from Kruschke book (Doing Bayesian...)
```{r}

library(dplyr) #data frame manipulation 
library(R2jags) # jags model
library(mcmcplots) # visualization of posterior plots
library(plyr) # data frame manipulation 
library(coda) # summarize models 


### Model block 

cat(file = "scripts/goose_norm_hb0_hyear_hlat_int_custom_fix.txt",
"
data {
xm <- mean(x)
latm <- mean(lat)
intm <- mean(int)
ym <- mean(y)
xsd <- sd(x)
latsd <- sd(lat)
intsd <- sd(int)
ysd <- sd(y)
}
model {
  for (i in 1:Ntotal) {
    zy[i] ~ dnorm(zbeta0[s[i]] + zbeta1[s[i]] * zx[i] + 
        zbeta2[s[i]] * zlat[i] + zbeta3[s[i]] * zint[i],  
    1/zsigma[s[i]]^2)
  }
  # 'hyper' parameters
  #   for normal distributions that species intercepts & slopes come from
  for (j in 1:Nsubj) {
    zbeta0[j] ~ dnorm(zbeta0mu, 1/(zbeta0sigma)^2)
    zbeta1[j] ~ dnorm(zbeta1mu, 1/(zbeta1sigma)^2)
    zsigma[j] ~ dunif( 1.0E-3, 1.0E+3 )
  }
  
  # setting latitude & interaction priors for geese 2,4,5, and 6

for(i in c(2,4,5,6)) {  
  zbeta2[i] ~ dnorm(zbeta2mu, 1/(zbeta2sigma)^2)
  zbeta3[i] ~ dnorm(zbeta3mu, 1/(zbeta3sigma)^2)
}

# special latitude priors for geese 1 and 3 with no variation in latitude 
for(i in c(1,3)) {
zbeta2[i] ~ dnorm(0, 1/(0.001)^2) # setting slope to be near zero
zbeta3[i] ~ dnorm(0, 1/(0.001)^2) # setting slope to be near zero  
}
 
  # priors on standardized scale:
  zbeta0mu ~ dnorm(0, 1/(10)^2)
  zbeta1mu ~ dnorm(0, 1/(10)^2)
  zbeta2mu ~ dnorm(0, 1/(10)^2)
  zbeta3mu ~ dnorm(0, 1/(10)^2)
  zbeta0sigma ~ dunif( 1.0E-3, 1.0E+3 )
  zbeta1sigma ~ dunif( 1.0E-3, 1.0E+3 )
  zbeta2sigma ~ dunif( 1.0E-3, 1.0E+3 )
  zbeta3sigma ~ dunif( 1.0E-3, 1.0E+3 )
  
  # Tranform to original non-standardized scale
  for ( j in 1:Nsubj ) {
    beta1[j] <- zbeta1[j] * ysd / xsd
    beta2[j] <- zbeta2[j] * ysd / latsd
    beta3[j] <- zbeta3[j] * ysd / intsd
    beta0[j] <- zbeta0[j] * ysd + ym - ((zbeta1[j] * xm * ysd / xsd) + (zbeta2[j] * latm * ysd / latsd) +
                (zbeta3[j] * intm * ysd / intsd))
    sigma[j] <- zsigma[j] * ysd
  }
  beta1mu <- zbeta1mu * ysd / xsd
  beta2mu <- zbeta2mu * ysd / latsd
  beta3mu <- zbeta3mu * ysd / intsd
  beta0mu <- zbeta0mu * ysd + ym - ((zbeta1mu * xm * ysd / xsd) + (zbeta2mu * latm * ysd / latsd) + (zbeta3mu * intm * ysd / intsd))
  
}
"
)


```
# specify simple model with only year
```{r}
### Model block 

cat(file = "scripts/goose_norm_hb0_hyear.txt",
"
data {
xm <- mean(x)
ym <- mean(y)
xsd <- sd(x)
ysd <- sd(y)
}

model {
  for (i in 1:Ntotal) {
    zy[i] ~ dnorm(zbeta0[s[i]] + zbeta1[s[i]] * zx[i],  
    1/zsigma[s[i]]^2)
  }
  
  # 'hyper' parameters
  #   for normal distributions that species intercepts & slopes come from
  for (j in 1:Nsubj) {
    zbeta0[j] ~ dnorm(zbeta0mu, 1/(zbeta0sigma)^2)
    zbeta1[j] ~ dnorm(zbeta1mu, 1/(zbeta1sigma)^2)
    zsigma[j] ~ dunif( 1.0E-3, 1.0E+3 )
  }
 
  # priors on standardized scale:
  zbeta0mu ~ dnorm(0, 1/(10)^2)
  zbeta1mu ~ dnorm(0, 1/(10)^2)
  zbeta0sigma ~ dunif( 1.0E-3, 1.0E+3 )
  zbeta1sigma ~ dunif( 1.0E-3, 1.0E+3 )
  
  # Tranform to original non-standardized scale
  for ( j in 1:Nsubj ) {
    beta1[j] <- zbeta1[j] * ysd / xsd
    beta0[j] <- zbeta0[j] * ysd + ym - zbeta1[j] * xm * ysd / xsd
    sigma[j] <- zsigma[j] * ysd
  }
  beta1mu <- zbeta1mu * ysd / xsd
  beta0mu <- zbeta0mu * ysd + ym - zbeta1mu * xm * ysd / xsd
  
}
"
)
```
# Run the interaction model
Key for interpreting model output:

1 = "Atlantic Brant"
2 = Black Brant
3 = Cackling Goose
4 = Greater White-fronted Goose
5 = Lesser Snow Goose
6 = Ross's Goose
```{r}
modfile_int <- "scripts/goose_norm_hb0_hyear_hlat_int_custom_fix.txt" 
modfile_noint <- "scripts/goose_norm_hb0_hyear.txt" 

# Run the model with latitude and interaction (or run without those by changing model.file name)
jags_mod <- jagsUI::jags(
  data = goose_jags_data,
  model.file = modfile_int,
  n.iter = 300000,
  n.adapt = 10000,
  parameters.to.save = c(
    "zbeta0",
    "zbeta1",
    "zbeta2",
    "zbeta3",
    "zbeta0mu",
    "zbeta1mu",    
    "zbeta2mu",    
    "zbeta3mu",
    "zbeta1sigma",
    "zbeta2sigma",
    "zbeta3sigma",
    "zsigma",
    "beta0",
    "beta1",
    "beta2",
    "beta3",
    "sigma",
    "beta0mu",
    "beta1mu",
    "beta2mu",
    "beta3mu",
    "beta0u",
    "beta1u",
    "beta2u",
    "beta3u",
    "sigmau",
    "beta0muu",
    "beta1muu",
    "beta2muu",
    "beta3muu"
  ),
  # inits = list(inits, inits, inits),
  n.chains = 3,
  parallel = T,
  n.thin = 3
)

saveRDS(jags_mod,
        "output/goose_norm_hb0_hyear_hlat_int_1_3_zerolat_fix_bigrun_SC.rds")

# saveRDS(jags_mod,
#         "output/goose_norm_hb0_hyear_hlat_int_1_3_zerolat_fix.rds")
# saveRDS(jags_mod,
#         "output/goose_norm_hb0_hyear.rds")

# view basic output of models
jags_mod

# examine traceplots to assess model convergence
jagsUI::traceplot(jags_mod)

```
# Plot goose response - with interaction
```{r}
# load model
mod <- 
  readRDS("output/goose_norm_hb0_hyear_hlat_int_1_3_zerolat_fix_bigrun_SC.rds")

# make a dataframe of goose names ###MAKE SURE PLYR NOT ATTACHED
geese_names <- f %>%
  group_by(species) %>%
  summarize(minlat = min(lat),
            maxlat = max(lat),
            zminlat = min(zlat),
            zmaxlat = max(zlat)) %>%
  mutate(num = 1:goose_jags_data$Nsubj,
         ztrend = round(mod$mean$beta1, 5))

# make a dataframe of years
minyr <- min(f$year)
maxyr <- max(f$year)
yearnum <- data.frame(yearnums = seq(minyr:maxyr),
                      years = minyr:maxyr,
                      zyears = sort(unique(goose_jags_data$x)))

# create 2 empty lists to put posterior samples into
gooseminlatlist <- list()
goosemaxlatlist <- list()
ysminlatlist <- list()
ysmaxlatlist <- list()

# for each goose
for (g in 1:goose_jags_data$Nsubj) {
  # get posterior samples of y predictions for each year
  for (y in 1:30) {
    # put the samples into elements of a list by year
    ysminlatlist[[y]] <- mod$sims.list$beta0[, g] +
      mod$sims.list$beta1[, g] * yearnum$years[y] +
      mod$sims.list$beta2[, g] * geese_names$minlat[g] +
      mod$sims.list$beta3[, g] * geese_names$minlat[g] * yearnum$years[y]
    
    ysmaxlatlist[[y]] <- mod$sims.list$beta0[, g] +
      mod$sims.list$beta1[, g] * yearnum$years[y] +
      mod$sims.list$beta2[, g] * geese_names$maxlat[g] +
      mod$sims.list$beta3[, g] * geese_names$maxlat[g] * yearnum$years[y]

  }
  
  # put the list of predictions into a big list with all geese
  gooseminlatlist[[g]] <- ysminlatlist
  goosemaxlatlist[[g]] <- ysmaxlatlist
  
}

# make an empty list to put summarized posteriors into
goosesum.minlat_list <- list()
goosesum.maxlat_list <- list()

# summarize the posteriors of predicted y for each species and year
for(i in 1:goose_jags_data$Nsubj){
# for each goose
# first make a dataframe of goose name and year
  # then bind it to posterior quantiles (min latitude)
goosesum.minlat_list[[i]] <- 
  cbind(data.frame(name = geese_names$species[i],
                 year = minyr:maxyr,
                 lat = "min"),
      data.frame(
do.call(rbind,
lapply(gooseminlatlist[[i]], 
       FUN = function(x) quantile(x, c(0.025,0.10, 0.5, 0.9, 0.975)))
)
)
)

# max lat
goosesum.maxlat_list[[i]] <- 
  cbind(data.frame(name = geese_names$species[i],
                 year = minyr:maxyr,
                 lat = "max"),
      data.frame(
do.call(rbind,
lapply(goosemaxlatlist[[i]], 
       FUN = function(x) quantile(x, c(0.025,0.10, 0.5, 0.9, 0.975)))
)
)
)

}

meany <- mean(f$dist)
sdy <- sd(f$dist)
sdx <- sd(f$year)

# collapse the list of posterior summaries into a df for plotting
goosesum <- do.call(rbind, goosesum.minlat_list) %>%
  bind_rows(do.call(rbind, goosesum.maxlat_list)) %>%
  rename(q2.5 = 4, q10 = 5, med = 6, q90 = 7, q97.5 = 8) %>%
  # mutate(q2.5 = meany+(q2.5*sdy),
  #        med = meany+(med*sdy),
  #        q97.5 = meany+(q97.5*sdy),
  #        q10 = meany+(q10*sdy),
  #        q90 = meany+(q90*sdy)) %>%
  filter(!(name == "Cackling Goose" & lat == "max"),
         !(name == "Atlantic Brant" & lat == "max"))

# plot the predictions

# make dataframe for trend annotations
maxdist <- goosesum %>% group_by(name) %>% summarize(max = max(q97.5)+100)
ann <- geese_names %>% 
  left_join(maxdist, by = c("species" = "name")) %>%
  mutate(trend = ztrend * sdy / sdx,
    trend2 = paste0(round(trend,2), " km/yr")) %>%
  rename(name = species)

goosesum %>%
  ggplot() +
  geom_ribbon(aes(x = year, ymin = q2.5, 
                  ymax = q97.5, fill = lat, group = lat),
              fill = "gray", alpha = 0.6) +
  geom_ribbon(aes(x = year, ymin = q10, 
                  ymax = q90, group = lat, fill = lat),
              fill = "gray40", alpha = 0.6) +
  geom_line(aes(x = year, y = med, linetype = lat),
            color = "black", linewidth = .7) +
  # geom_text(aes(x = 2017, y = max, 
  #                      group = name, label = trend2),
  #           data = ann, hjust = 1, vjust = 1,
  #           size = 3) +
  facet_wrap(~name, scales = "free_y") +
  scale_x_continuous(breaks = c(1,11,21,30), labels = c(1990, 2000, 2010, 2019)) +
  labs(x = "Year of banding", 
       y = "Mean distance between \n breeding and winter (km)") +
  theme_linedraw() +
 theme(strip.background = element_blank(),strip.text = element_text(hjust = 0,color="black")) +
  guides(linetype = "none") +
    guides(linetype = "none")+scale_y_continuous(labels = comma)

# ggsave("goose_trends_lat_int2.png", height = 6, width = 9, dpi = 400)
  
```
# goose parameter table
```{r}
# load model
mod <- 
  readRDS("output/goose_norm_hb0_hyear_hlat_int_1_3_zerolat_fix_bigrun.rds")

minmax <- f %>% 
  group_by(species) %>% 
  summarize(min = min(lat), 
            mean = mean(lat),
            max = max(lat))

# get predicted trends at different latitudes

trendminlist <- list()
trendmeanlist <- list()
trendmaxlist <- list()
for(i in 1:nrow(mod$sims.list$beta3)){
trendminlist[[i]] <- mod$sims.list$beta1[i,] + mod$sims.list$beta3[i,]*minmax$min
trendmeanlist[[i]] <- mod$sims.list$beta1[i,] + mod$sims.list$beta3[i,]*minmax$mean
trendmaxlist[[i]] <- mod$sims.list$beta1[i,] + mod$sims.list$beta3[i,]*minmax$max
}
trendmin_cl <- apply(do.call(rbind, trendminlist), 2, FUN = function(x) quantile(x, c(0.025,0.975)))
trendmean_cl <- apply(do.call(rbind, trendmeanlist), 2, FUN = function(x) quantile(x, c(0.025,0.975)))
trendmax_cl <- apply(do.call(rbind, trendmaxlist), 2, FUN = function(x) quantile(x, c(0.025,0.975)))
trendmin_m <- apply(do.call(rbind, trendminlist), 2, mean)
trendmean_m <- apply(do.call(rbind, trendmeanlist), 2, mean)
trendmax_m <- apply(do.call(rbind, trendmaxlist), 2, mean)

# format posterior means for table
means <- data.frame(
sp = sort(unique(f$species)),
int = mod$mean$beta0,
year = mod$mean$beta1,
lat = mod$mean$beta2,
year_x_lat = mod$mean$beta3,
sigma = mod$mean$sigma,
predmin = trendmin_m,
predmean = trendmean_m,
predmax = trendmax_m
) %>%
  t()

lcls <- data.frame(
sp = sort(unique(f$species)),
int = mod$q2.5$beta0,
year = mod$q2.5$beta1,
lat = mod$q2.5$beta2,
year_x_lat = mod$q2.5$beta3,
sigma = mod$q2.5$sigma,
predmin = trendmin_cl[1,],
predmean = trendmean_cl[1,],
predmax = trendmax_cl[1,]
) %>%
  t()

ucls <- data.frame(
sp = sort(unique(f$species)),
int = mod$q97.5$beta0,
year = mod$q97.5$beta1,
lat = mod$q97.5$beta2,
year_x_lat = mod$q97.5$beta3,
sigma = mod$q97.5$sigma,
predmin = trendmin_cl[2,],
predmean = trendmean_cl[2,],
predmax = trendmax_cl[2,]
) %>%
  t()

tab = data.frame(labs = c("species", "Intercept", "Year", "Latitude", "Year x Latitude", "Sigma",
                    "Predicted trend at min. latitude (km/yr)", 
                    "Predicted trend at mean latitude (km/yr)", 
                    "Predicted trend at max. latitude (km/yr)"),
          means = c(means[,1], means[,2], means[,3], means[,4], means[,5], means[,6]),
           lcls = c(lcls[,1], lcls[,2], lcls[,3], lcls[,4], lcls[,5], lcls[,6]),
           ucls = c(ucls[,1], ucls[,2], ucls[,3], ucls[,4], ucls[,5], ucls[,6]))

library(kableExtra)

tab_table <- tab %>%
  kbl(caption = "parameters") %>%
  kable_classic(full_width = F, html_font = "Cambria")

tab_table

# save the table to HTML file
save_kable(tab_table, "Table_model_results.html")

```
# goose change stats
```{r}

# strongest decreases
test = goosesum %>% filter(lat == "min")

(t.start <- test[test$year=="1",]$med)
(t.end <- test[test$year=="30",]$med)

t.start - t.end
# 

(t.start - t.end) / length(unique(goosesum$year))

unique(test$name)

# strongest increases
test = goosesum %>% filter(lat == "max")

(t.start <- test[test$year=="1",]$med)
(t.end <- test[test$year=="30",]$med)

t.end - t.start
# 152.832928 151.763766  79.307081   2.706668

(t.end - t.start) / length(unique(goosesum$year))

unique(test$name)
# [1] "Atlantic Brant"              "Black Brant"                 "Cackling Goose"              "Greater White-fronted Goose"
# [5] "Lesser Snow Goose"           "Ross's Goose"
```

